# Finding Data {#finding-data}

```{r finding-data-options, include=FALSE}
library(knitr)
opts_chunk$set(results='hide', warning=FALSE, message=FALSE)
```

```{r finding-data-libraries}
# Bioconductor
library(GEOquery)

# Domain specific
library(rentrez)

# General purpose
library(curl)
library(httr)
library(urltools)
library(xml2)

# Tidyverse
library(fs)
library(lubridate)
library(janitor)
library(tidyverse)
```

## Managing downloads
Let's put all of our downloaded data into a single directory.

```{r create-data-dir}
data_dir <- dir_create("data")
```

## Exploring ImmuneSigDB

What's in ImmuneSigDB? To find out, [register](http://software.broadinstitute.org/gsea/register.jsp) with GSEA and download the "Current MSigDB xml file"; at the time of this writing, that's `msigdb_v6.2.xml`.

```{r download-msigdb}
download_dir <- fs::path(data_dir, "msigdb_files")
msigdb_file <- fs::path(download_dir, "msigdb.xml")

if (!dir_exists(download_dir)) {
  dir_create(download_dir)
  login_url <- "http://software.broadinstitute.org/gsea/j_spring_security_check"
  login_info <- list(j_username = "jeff.hammerbacher@gmail.com", j_password = "password")
  msigdb_url <- "http://software.broadinstitute.org/gsea/msigdb/download_file.jsp?filePath=/resources/msigdb/6.2/msigdb_v6.2.xml"
  POST(login_url, body = login_info, encode = "form")
  GET(msigdb_url, write_disk(msigdb_file), progress())
}
```

We are only interested in the C7 gene set, also known as ImmuneSigDB. For now we are using XPath to extract the relevant data; perhaps we should show how to explore the XML file and construct this query.

```{r extract-c7, dependson='create-data-dir'}
x <- read_xml(msigdb_file)
c7_xml <- xml_find_all(x, "//GENESET[contains(@CATEGORY_CODE,'C7')]")
c7_tib <- c7_xml %>% as_list %>% purrr::map(~ as_tibble(attributes(.x))) %>% bind_rows %>% clean_names
```