# Finding Data {#finding-data}

```{r finding-data-options, include=FALSE}
library(knitr)
opts_chunk$set(results='hide', warning=FALSE, message=FALSE)
```

```{r finding-data-libraries}
# Publication-related
library(kableExtra)
options(kableExtra.html.bsTable = T)

# Bioconductor
library(GEOquery)

# Domain specific
library(rentrez)

# General purpose
library(httr)
library(xml2)

# Tidyverse
library(fs)
library(glue)
library(lubridate)
library(janitor)
library(tidyverse)
```

## Managing downloads
Let's put all of our downloaded data into a single directory.

```{r finding-data-create-data-dir}
data_dir <- dir_create("data")
```

## ImmuneSigDB

What's in ImmuneSigDB? To find out, [register](http://software.broadinstitute.org/gsea/register.jsp) with GSEA and download the "Current MSigDB xml file"; at the time of this writing, that's `msigdb_v6.2.xml`.

```{r download-msigdb}
download_dir <- fs::path(data_dir, "msigdb_files")
msigdb_file <- fs::path(download_dir, "msigdb.xml")

if (!dir_exists(download_dir)) {
  dir_create(download_dir)
  login_url <- "http://software.broadinstitute.org/gsea/j_spring_security_check"
  login_info <- list(j_username = "jeff.hammerbacher@gmail.com", j_password = "password")
  msigdb_url <- "http://software.broadinstitute.org/gsea/msigdb/download_file.jsp?filePath=/resources/msigdb/6.2/msigdb_v6.2.xml"
  POST(login_url, body = login_info, encode = "form")
  GET(msigdb_url, write_disk(msigdb_file), progress())
}
```

We are only interested in the C7 gene set, also known as ImmuneSigDB. For now we are using XPath to extract the relevant data; perhaps we should show how to explore the XML file and construct this query.

```{r extract-c7}
x <- read_xml(msigdb_file)
c7_xml <- xml_find_all(x, "//GENESET[contains(@CATEGORY_CODE,'C7')]")
c7_tib <- c7_xml %>% as_list %>% purrr::map(~ as_tibble(attributes(.x))) %>% bind_rows %>% clean_names
```

Each row in `c7_tib` represents a gene set from ImmuneSigDB. We filter down to those gene sets we suspect relate to T cells.

```{r filter-tcells}
tcell_terms <- c("TCELL", "CD3", "CD4", "CD8")
c7_tcells <- c7_tib %>% rowwise %>% filter(any(str_detect(standard_name, tcell_terms))) %>% ungroup
```

We can now explore publications and GEO data sets used to build these gene sets. We are specifically interested in human, not mouse, T cells in this chapter. 

```{r tcell-pubs}
tcell_pubs <- c7_tcells %>%
  filter(organism == "Homo sapiens") %>%
  group_by(pmid) %>%
  dplyr::summarize(n_genesets = n(), n_geoids = n_distinct(geoid), geoids = toString(unique(geoid))) %>%
  mutate(n_geoids = if_else(geoids == "", 0L, n_geoids)) %>%
  arrange(desc(n_genesets))
```

Let's get some metdata for these papers. 

```{r pub-metadata}
pub_meta_esum <- entrez_summary(db = "pubmed", id = tcell_pubs$pmid, config = httr::config(http_version = 2))
pub_meta_list <- extract_from_esummary(pub_meta_esum, c("uid", "pubdate", "title")) %>% array_branch(2)
pub_meta_vec <- pub_meta_list %>% purrr::map(unlist) %>% flatten_chr()
pub_meta <- tribble(~pmid, ~pubdate, ~title, !!!pub_meta_vec)
pub_meta <- pub_meta %>% mutate(pubdate = parse_date_time(pubdate, orders = c("Y", "Ym", "Ymd")))
tcell_pubs <- tcell_pubs %>% left_join(pub_meta, by = "pmid")
```

```{r tcell-pubs-table, fig.cap = "T cell publications", echo = FALSE, results = "markup"}
tcell_pubs %>%
  mutate(
    pmid = text_spec(pmid, link = paste0("https://www.ncbi.nlm.nih.gov/pubmed/", pmid))
  ) %>%
  kable(escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  column_spec(5, width = "10em") %>%
  column_spec(6, width = "100em")
```

We're most interested in the expression profile of common T cell subsets in healthy subjects without an intervention.

The best candidates based on manual inspection of this table:

- [Immune response in silico (IRIS): immune-specific genes identified from a compendium of microarray expression data](https://www.ncbi.nlm.nih.gov/pubmed/15789058) (2005): [GSE22886](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE22886) has CD8, CD4, and memory T cells from 3 or 4 individuals profiled with the Affy U133A and B arrays. The paper that launched a thousand deconvolution papers. 
- [Deconvolution of blood microarray data identifies cellular activation patterns in systemic lupus erythematosus](https://www.ncbi.nlm.nih.gov/pubmed/19568420) (2009): [GSE11057](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE11057) seems to contain PBMC, Naive, EM, and CM expression profiles for 4 individuals profiled with the Affy U133 Plus 2.0 Array. I can't tell if these are from SLE patients or healthy subjects though. This data was re-analyzed in [An expression atlas of human primary cells: inference of gene function from coexpression networks](https://www.ncbi.nlm.nih.gov/pubmed/24053356) (2013) and in the MCP-counter paper [Estimating the population abundance of tissue-infiltrating immune and stromal cell populations using gene expression](https://www.ncbi.nlm.nih.gov/pubmed/27765066) (2016).

Other interesting papers:
- [Gene expression profiles during human CD4+ T cell differentiation](https://www.ncbi.nlm.nih.gov/pubmed/15210650) (2004): intrathymic T progenitor cells (ITTP), "double positive" thymocytes (DP), "single positive" thymocytes (SP4), naïve T cells from cord blood (CB4+), and naïve T cells from adult blood (AB4+) for 3 individuals profiled with Affy U133A and B arrays. Cool data for T cell development.
- [Functional heterogeneity of human effector CD8+ T cells](https://www.ncbi.nlm.nih.gov/pubmed/22174157) (2012): CXCR1+ and CXCR1- subsets of human effector CD27-CD28-CD8+ T cells 5 individuals profiled with the Affy U133 Plus 2.0 Array. Could easily pool data from individuals to get a nice CD8+ Teff profile.

## MCP-counter

The MCP-counter paper [Estimating the population abundance of tissue-infiltrating immune and stromal cell populations using gene expression](https://www.ncbi.nlm.nih.gov/pubmed/27765066) (2016) organizes expression data by cell type in their supplemental material.

I've combined data for T cells across tables S1, S4, and S5 in [my curated version](https://docs.google.com/spreadsheets/d/1AI00Q3MK53dZ3K1YqrncKoQm0Kxj1CFOQgfrl3lp7fM/edit). They've catalogued data for 1,776 samples from 64 GEO series (not on GEO, but included: data from [@Chtanova2005-zq]) for 22 different T cell phenotypes across 4 (?) different array platforms. We should explore this data!

```{r mcp-counter-data}
mcp_datasets <- c("GSE10586", "GSE11057", "GSE11188", "GSE11292", 
"GSE12875", "GSE13017", "GSE13906", "GSE14908", "GSE14924", "GSE14926", 
"GSE15659", "GSE17354", "GSE17922", "GSE18893", "GSE19069", "GSE20198", 
"GSE22045", "GSE22886", "GSE23293", "GSE23294", "GSE23295", "GSE23332", 
"GSE25087", "GSE26928", "GSE27291", "GSE2770", "GSE28107", "GSE28490", 
"GSE28491", "GSE29583", "GSE32901", "GSE32959", "GSE33670", "GSE36476", 
"GSE36769", "GSE37213", "GSE41909", "GSE42569", "GSE43177", "GSE43260", 
"GSE44392", "GSE44460", "GSE44621", "GSE45535", "GSE473", "GSE48558", 
"GSE49703", "GSE49877", "GSE50175", "GSE51540", "GSE52129", "GSE53455", 
"GSE53514", "GSE5580", "GSE56035", "GSE58867", "GSE60234", "GSE60235", 
"GSE6338", "GSE64914", "GSE6566", "GSE6740", "GSE7497", "GSE8835")

mcp_phenotypes <- c("Activated CD4+ T cells", "Activated Memory CD4 T cells", "Activated T cells", 
"Canonical CD4 Treg cells", "CD4 T cells", "CD8 T cells", "Central Memory CD4 T cells", 
"CXCR5+ Memory CD4 T cells", "Effector Memory CD4 T cells", "Effector Memory CD8 T cells", 
"Effector Memory T cells", "Memory CD4 T cells", "Naive CD4 T cells", 
"Naive T cells", "Resting Memory CD4 T cells", "T cells", "T gamma delta", 
"Th1", "Th17", "Th1Th17", "Th2", "Type 1 regulatory cells")
```

We can use `rentrez` to query the [GEO DataSets (GDS)](https://www.ncbi.nlm.nih.gov/gds/) database.
TODO(hammer): get the PMIDs associated with these GSEs, then make a table to compare to `tcell_pubs`.

```{r mcp-geo}
mcp_str <- str_c(mcp_datasets, collapse = "[ACCN] OR ")
query <- glue("({mcp_str}[ACCN]) AND GSE[ETYP]")
mcp_ids <- entrez_search(db = "gds", term = query, retmax = 100, config = httr::config(http_version = 2))
mcp_esum <- entrez_summary(db = "gds", id = mcp_ids$ids, config = httr::config(http_version = 2))
mcp_list <- extract_from_esummary(mcp_esum, c("gse", "n_samples", "title", "summary")) %>% array_branch(2)
mcp_vec <- mcp_list %>% purrr::map(unlist) %>% flatten_chr()
mcp <- tribble(~gse, ~n_samples, ~title, ~summary, !!!mcp_vec)
```

```{r mcp-table, fig.cap = "MCP GEO DataSets", echo = FALSE, results = "markup"}
mcp %>%
  arrange(desc(parse_integer(n_samples))) %>%
  kable(escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

