# Microarrays {#microarrays}

```{r global-options, include=FALSE}
library(knitr)
opts_chunk$set(results='hide', warning=FALSE, message=FALSE)
```

```{r libraries}
library(GEOquery)
library(oligo)
library(limma)
library(biobroom)

library(curl)
library(fs)
library(urltools)
library(xml2)

library(lubridate)
library(janitor)
library(tidyverse)
```

## Exploring ImmuneSigDB

What's in ImmuneSigDB? To find out, head over to the [GSEA downloads page](http://software.broadinstitute.org/gsea/downloads.jsp) (you'll need to register your email address first) and grab the "Current MSigDB xml file"; at the time of this writing, that's `msigdb_v6.2.xml`.

Let's put all of our downloaded data into a single directory.

```{r create-data-dir}
data_dir <- dir_create("data")
```

Put the MSigDB file into this directory at `msigdb_files/msigdb_v6.2.xml`. We are only interested in the C7 gene set, also known as ImmuneSigDB.

```{r munge-msigdb-xml, dependson='create-data-dir'}
x <- read_xml(fs::path(data_dir, "msigdb_files/msigdb_v6.2.xml"))
c7 <- xml_find_all(x, "//GENESET[contains(@CATEGORY_CODE,'C7')]")
c7_tib <- c7 %>% as_list %>% purrr::map(~ as_tibble(attributes(.x))) %>% bind_rows %>% clean_names
```

Here are a few queries of our cleaned data to find T cell datasets.

```{r query-msigdb-xml, dependson='munge-msigdb-xml'}
c7_tib %>% filter(str_detect(standard_name, 'TCELL') | str_detect(standard_name, 'CD3') | str_detect(standard_name, 'CD4') | str_detect(standard_name, 'CD8')) %>% group_by(organism) %>% count() # Homo sapiens 612, Mus musculus 1124
# Human T cell publications
c7_tib %>% filter(str_detect(standard_name, 'TCELL') | str_detect(standard_name, 'CD3') | str_detect(standard_name, 'CD4') | str_detect(standard_name, 'CD8')) %>% filter(organism == 'Homo sapiens') %>% group_by(pmid) %>% count() %>% arrange(desc(n))
```

## Reproducing a single gene set from ImmuneSigDB
Let's just get the CEL files used to generate the [GSE2770_IL12_AND_TGFB_ACT_VS_ACT_CD4_TCELL_6H_DN](http://software.broadinstitute.org/gsea/msigdb/geneset_page.jsp?geneSetName=GSE2770_IL12_AND_TGFB_ACT_VS_ACT_CD4_TCELL_6H_DN) gene set from ImmuneSigDB [@Godec2016-ku].

First, we download data for GSE2770 from [GEO](https://www.ncbi.nlm.nih.gov/geo/). There are 3 elements in this list, 1 for each platform.

```{r getgeo-gse2770, dependson='create-data-dir'}
download_dir <- fs::path(data_dir, "series_matrix_files")
if (!dir_exists(download_dir)) {
  dir_create(download_dir)
  GSE2770 <- getGEO("GSE2770", getGPL=FALSE, destdir=download_dir)
} else {
  GSE2770 <- dir_ls(download_dir, glob="*series_matrix*") %>%
    purrr::map(~ getGEO(filename=.x, getGPL=FALSE))
}
```

Next, we pull out the phenotype data.

```{r pgse2770, dependson='getgeo-gse2770'}
pGSE2770 <- GSE2770 %>%
  purrr::map(~ as(pData(.x), "data.frame")) %>%
  bind_rows() %>%
  as_tibble()
```

Then we tidy up the phenotype data.

```{r pgse2770tidy, dependson='pgse2770'}
pGSE2770tidy <- pGSE2770 %>%
  select(platform_id, geo_accession, supplementary_file, title) %>%
  separate(title, c("cells", "treatment", "time", "replicate", "platform"), sep="[_\\(\\)]") %>%
  select(-cells) %>%
  mutate(replicate = map_chr(replicate, ~ str_split(.x, boundary("word"))[[1]][2])) %>%
  mutate(time = as.duration(time))
```

Here we get the CEL file URLs for the two conditions used to generate this particular gene set.

```{r get-cel-list, dependson='pgse2770tidy'}
# get just the U133A cel files
cel_file_urls <- pGSE2770tidy %>%
  filter(time == duration("6h") & (treatment %in% c("antiCD3+antiCD28+IL12+TGFbeta", "antiCD3+antiCD28")) & platform == "U133A") %>%
  .$supplementary_file
```

We then download the CEL files.

```{r download-cels, dependson='get-cel-list'}
download_dir <- fs::path(data_dir, "cel_files")
if (!dir_exists(download_dir)) {
  dir_create(download_dir)
  cel_file_names <- path_file(url_parse(cel_file_urls)$path)
  cel_files_local <- map2(cel_file_urls, fs::path(download_dir, cel_file_names), ~ curl_download(.x, .y)) %>% simplify()
} else {
  cel_files_local <- dir_ls(download_dir, glob="*.cel*", ignore.case=TRUE)
}
```

And read them into an oligo ExpressionFeatureSet object. Note the feature data (i.e. the probeset IDs) are not stored in the feature data of the ExpressionSet object, but rather in a SQLlite database [pd.hg.u133a](https://bioconductor.org/packages/release/data/annotation/html/pd.hg.u133a.html) somewhere on disk that the call to `rma()` will pick up.

```{r read-cels, dependson='download-cels'}
raw_data <- read.celfiles(cel_files_local)
```

Run RMA on our batch of CEL files to be compared.

```{r rma, dependson='read-cels'}
processed_data <- rma(raw_data)
as_tibble(exprs(processed_data), rownames = "probe_set_id")
```

Because we want to do our differential expression analysis at the gene level, not the probeset level, we need to map probeset IDs to gene symbols. To ensure comparability with the GSEA gene sets, we'll use the annotations provided by the Broad that are used in GSEA.

```{r download-chip}
chip_file_url <- "ftp://ftp.broadinstitute.org/pub/gsea/annotations/HG_U133A.chip"
chip_file_name <- path_file(url_parse(chip_file_url)$path)

download_dir <- fs::path(data_dir, "chip_files")
if (!dir_exists(download_dir)) {
  dir_create(download_dir)
  chip_file <- curl_download(chip_file_url, fs::path(download_dir, chip_file_name))
} else {
  chip_file <- fs::path(download_dir, chip_file_name)
}
```

We need to clean up the data a bit: the import produced a bogus column (X4), and we only want one gene symbol per probeset ID.

```{r tidy-chip, dependson='download-chip'}
U133A_chip_raw <- read_tsv(chip_file)
U133A_chip <- U133A_chip_raw %>% 
  select(-X4) %>%
  clean_names() %>%
  rowwise() %>%
  mutate(first_gene_symbol = str_trim(str_split(gene_symbol, "///")[[1]][1])) 
```

Finally we can join with our processed data!

```{r join-chip, dependson=c('rma', 'download-chip')}
eset_tib <- as_tibble(exprs(processed_data), rownames = "probe_set_id")
eset_tib_genes <- eset_tib %>% left_join(U133A_chip, by = "probe_set_id")
```

We now remove probesets that don't map to a gene symbol; note chip files encode this fact with `---`, which we demonstrate first before the filter.

```{r remove-unmapped-probesets, dependson='join-chip'}
eset_tib_genes %>% filter(is.na(first_gene_symbol)) %>% count() # 0
eset_tib_genes %>% filter(first_gene_symbol == "---") %>% count() # 1,109
eset_tib_genes <- eset_tib_genes %>% filter(first_gene_symbol != "---")
```

Some genes have multiple probesets that map to them. GSEA handles this situation by only keeping the maximum intensity value across all probesets that map to the gene. In GSEA terminology, we are using the max probe algorithm to collapse our pobesets at the gene level.

```{r collapse-probesets, dependson='remove-unmapped-probesets'}
eset_tib_genes_only <- eset_tib_genes %>%
  select(-c(probe_set_id, gene_symbol, gene_title)) %>%
  group_by(first_gene_symbol) %>%
  summarise_all(max)
```

Now we'll take our new genes-only expression data and put it into a form expected by `limma`.

```{r make-expressionset, dependson='collapse-probesets'}
eset_genes <- eset_tib_genes_only %>%
  as.data.frame() %>%
  column_to_rownames(var="first_gene_symbol") %>%
  data.matrix() %>%
  ExpressionSet()
```

One last step: we need to make the design matrix for this comparison.

```{r make-design}
treatments <- factor(c(1,2,2,1), labels=c("untreated", "treated")) # weird ordering of files
design <- model.matrix(~treatments)
```

Finally we use `limma` to perform a differential expression analysis!

```{r do-dge, dependson=c('make-expressionset', 'make-design')}
fit_genes <- lmFit(eset_genes, design)
fit2_genes <- eBayes(fit_genes)
results_genes <- topTable(fit2_genes, number=Inf)
results_genes_tib <- as_tibble(results_genes, rownames = "gene")
```

Let's look at our results.

```{r volcano, dependson='do-dge', fig.cap = "Volcano plot made with biobroom", echo = FALSE}
ggplot(tidy(fit2_genes), aes(x=estimate, y=log(p.value), color=statistic)) + 
  geom_point() +
  theme_bw()
```

